---
title: "ANTsPyMM: Blind Image Quality Control (QC) Summary"
author: "Automated Analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    df_print: kable
---

```{r setup, echo=FALSE, include=FALSE, message=FALSE, warning=FALSE}
# --- R Environment Setup ---
# Load only the necessary libraries for this report.
# reticulate: The bridge between R and Python.
# DT: For creating interactive HTML data tables.
# pander: For creating clean, static markdown tables.
library(reticulate)
library(DT)
library(pander)

# Set global options for table formatting.
options(digits=3)
panderOptions('table.split.table', Inf) # Prevent pander from splitting wide tables
```

### 1. Introduction

This report demonstrates the `antspymm.blind_image_assessment` function for performing automated, no-reference Quality Control on medical images. We will first simulate a set of images with varying characteristics (noise, smoothness, etc.) and then run the QC analysis on them. The results are presented in interactive and static tables.

This script is designed to be fully portable and self-contained. It creates temporary files in a system-appropriate location and cleans up after itself upon completion.

### 2. Simulating Images for QC

First, we set up a Python environment and generate several test images. A temporary directory is created to store these images, ensuring the script does not leave files on the system.

```{python setup-and-simulate, message=FALSE, warning=FALSE}
# --- Python Environment and Image Simulation ---
import siq
import ants
import antspymm
import os
import tempfile
import shutil
import pandas as pd

# Create a single, system-agnostic temporary directory for all output files.
# This makes the script portable (works on Linux, macOS, Windows).
temp_dir = tempfile.mkdtemp()
print(f"Using temporary directory: {temp_dir}")

# List to hold the paths of all generated files for processing.
generated_files = []

# --- Image 1: Baseline Simulated Image ---
print("Simulating baseline image...")
eximg = siq.simulate_image()
exfn_base = os.path.join(temp_dir, 'baseline.nii.gz')
ants.image_write(eximg, exfn_base)
ants.plot( eximg )
generated_files.append(exfn_base)

# --- Image 2: Additive Gaussian Noise ---
print("Simulating image with Gaussian noise...")
img_gauss = siq.simulate_image().add_noise_to_image('additivegaussian', [0, 1])
exfn_gauss = os.path.join(temp_dir, 'noise_gaussian.nii.gz')
ants.image_write(img_gauss, exfn_gauss)
ants.plot( img_gauss )
generated_files.append(exfn_gauss)

# --- Image 3: Salt & Pepper Noise ---
print("Simulating image with Salt & Pepper noise...")
img_sp = siq.simulate_image()
img_sp = ants.add_noise_to_image(img_sp, 'saltandpepper', [0.2, -1, 1])
exfn_sp = os.path.join(temp_dir, 'noise_saltpepper.nii.gz')
ants.image_write(img_sp, exfn_sp)
ants.plot( img_sp )
generated_files.append(exfn_sp)

# --- Image 4: Smoothed Image ---
print("Simulating a smoothed image...")
img_smooth = siq.simulate_image().smooth_image(3)
exfn_smooth = os.path.join(temp_dir, 'smooth.nii.gz')
ants.image_write(img_smooth, exfn_smooth)
ants.plot( img_smooth )
generated_files.append(exfn_smooth)

# --- Image 5: Time Series Image ---
print("Simulating a 4D time-series image...")
img_ts = siq.simulate_image([16, 16, 16, 4])
exfn_ts = os.path.join(temp_dir, 'timeseries.nii.gz')
ants.image_write(img_ts, exfn_ts)
generated_files.append(exfn_ts)

print(f"\nGenerated {len(generated_files)} files for analysis.")
```

### 3. Running Batch QC Analysis

Now we loop through the generated images and run `blind_image_assessment` on each. The results are collected into a single Pandas DataFrame.

```{python run-qc, message=FALSE, warning=FALSE}
# --- Batch QC Processing ---
# A more efficient way to build a DataFrame is to append dictionaries to a
# list and then call the DataFrame constructor once.
qc_results = []
for f in generated_files:
    print(f"Running QC on: {os.path.basename(f)}...")
    myqc = antspymm.blind_image_assessment(f)
    qc_results.append(myqc)

# Concatenate all results into a single DataFrame.
qcdf = pd.concat(qc_results, axis=0)

# The 'qcdf' DataFrame is now automatically available to R via reticulate as py$qcdf
# No need to write to a temporary CSV file.
```

### 4. QC Results

The following tables display the QC metrics for all simulated images.

#### 4.1. Interactive QC Data Table

This table is sortable and searchable, which is useful for exploring the results dynamically.

```{r interactive-table, echo=FALSE}
# Access the Python DataFrame 'qcdf' directly using the reticulate 'py' object.
# This avoids the need for intermediate files and is much more efficient.
qc_data <- py$qcdf

# Use DT::datatable for a rich, interactive HTML table.
DT::datatable(
  qc_data,
  caption = 'Interactive table of blind QC metrics for simulated images.',
  options = list(pageLength = 10, scrollX = TRUE)
)
```

#### 4.2. Static QC Data Table

This is a clean, static view of the same data, suitable for printing or including in a non-interactive document.

```{r static-table, echo=FALSE, results='asis'}
# Use pander for a well-formatted, static markdown table.
pander(qc_data, caption = 'Static table of blind QC metrics.')
```

### 5. QC Metric Definitions

The metrics reported in the tables above are defined as follows:

`fn`
: The filename of the processed image.

`noise`
: Estimated noise level, calculated as the mean absolute difference between the original image and a smoothed version of it.

`snr`
: **Signal-to-Noise Ratio**. Defined from automatically generated foreground (fg) and background (bg) masks as: `mean(fg) / std(bg)`.

`cnr`
: **Contrast-to-Noise Ratio**. Defined as: `(mean(fg) - mean(bg)) / std(bg)`.

`psnr`
: **Peak Signal-to-Noise Ratio**. Calculated between the original and a smoothed version of the image.

`ssim`
: **Structural Similarity**. Measures the similarity between the original and a smoothed version of the image. Higher values indicate greater similarity.

`mi`
: **Mutual Information**. Measures the statistical dependency between the original and smoothed image.

`reflection_error`
: Mean absolute error between the original image and a version reflected across the mid-sagittal plane. Useful for detecting severe spatial artifacts or asymmetries.

`EVR`
: **Eigenvalue Ratio**. A measure of image texture and complexity. Values near 0 suggest very smooth images, while values near 1 suggest pure noise. Typical medical images fall in the 0.6-0.7 range.

`msk_vol`
: The volume of the automatically detected foreground mask, in cubic millimeters.

`spc*`
: The voxel spacing (e.g., `spcx`, `spcy`, `spcz`).

`org*`
: The image origin coordinates.

`dim*`
: The image dimensions (number of voxels along each axis).

`slice`
: The slice index, primarily used for 4D time-series data.

`modality`
: Inferred image modality, typically based on the filename.

`mriseries`, `mrimfg`, `mrimodel`
: Metadata fields (series description, manufacturer, model) read from a sidecar JSON file, if available.

```{python cleanup, include=FALSE, echo=FALSE}
# --- Cleanup ---
# This final chunk runs invisibly to remove the temporary directory and all
# its contents, ensuring the script leaves the system clean.
try:
    print(f"Cleaning up temporary directory: {temp_dir}")
    shutil.rmtree(temp_dir)
    print("Cleanup successful.")
except NameError:
    print("No temporary directory to clean up.")
```
